Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-10 paper/manuscript.tex
full draft of methods

\begin{figure}
\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.68\textwidth]{figures/entrenchment.pdf}}
\caption{\label{fig:entrenchment}
Entrenchment of substitutions during Env evolution.
There are 12,521 possible amino-acid mutations at the 659 mutagenized sites alignable between BG505 and BF520.
The blue densities show the effects of all these mutations to each Env.
The orange densities show the effects of just the 92 mutations that convert BG505 to BF520 or vice versa.
In the absence of entrenchment, mutating a site in BG505 to its identity in BF520 should have the opposite effect of mutating the site in BF520 to its identity in BG505.
In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution -- and both distributions should be centered around zero if the two Envs are equivalently functional.
Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions are often entrenched in the Env in which they have fixed.
The effect of a mutation is quantified as the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
}
\end{figure}

@@

The end primers for the BG505 mutagenesis were: 5'-tgaaggcaaaactactggtccgtctcgagcagaagacagtggcaatgaga-3' and 5'-gctacaaatgcatataacagcgtctcattctttccctaacctcaggcca-3'.
As with BF520, we cloned the BG505 \textit{env} libraries into the \textit{env} locus of the full-length proviral genome of HIV strain Q23 \citep[another subtype-A transmitted/founder virus;][]{poss1999variants} using the high-efficiency cloning vector described in \citet{dingens2017comprehensive}.
For this cloning, we digested the cloning vector with BsmBI, and then used PCR to elongate the amplicons to include 30 nucleotides at each end that were identical in sequence to the ends of the BsmBI-digested vector.
The primers for this PCR were:
5'-agataggttaattgagagaataagagaaagagcagaagacagtggcaatgagagtgatgg-3' and 5'-ctcctggtgctgctggaggggcacgtctcattctttccctaacctcaggccatcc-3'.
Next, we used NEBuilder HiFi DNA Assembly (NEB, E2621S) to clone the \textit{env} amplicons into the BsmBI-digested plasmids.
We purified the assembled products using Agencourt AMPure XP beads (Beckman Coulter, A63880) using a bead-to-sample ratio of 1.5, and then transformed the purified products into Stellar electrocompetent cells (Takara, 636765).
The transformations yielded between 1.5-3.6 million unique clones for each of the three replicate libraries, as estimated by plating 1:2,000 dilutions of the transformations.
We scraped the plated colonies and maxiprepped the plasmid DNA; unlike in \citet{dingens2017comprehensive}, we did not include a 4-hour outgrowth step after the scraping step.
For the wildtype controls, we maxiprepped three independent cultures of wildtype BG505 \textit{env} cloned into the same Q23 proviral plasmid.
See Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger} and Figure~\ref{fig:mutfreqs}A for information on the average mutation rate in these libraries as estimated by Sanger sequencing and deep sequencing, respectively.

@@

\subsection{Analysis of deep-sequencing data}
We analyzed the deep-sequencing data using the \texttt{dms\_tools2} software package~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
A Jupyter notebook that performs the entire analysis including generation of most of the figures in this paper is in Supplementary~file~\ref{suppfile:code}, which also contains all necessary input data to fully reproduce the analysis.
An HTML rendering of this notebook is in Supplementary~file~\ref{suppfile:html}.
Citations are also due to \texttt{weblogo}~\citep{crooks2004weblogo} and \texttt{ggseqlogo}~\citep{?}, which were used in the generation of the logoplots. 
\jdbcomment{also joypy?}

The Jupyter notebooks in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} also contain numerous plots that summarize relevant aspects of the deep sequencing such as read depth, per-codon mutation frequency, mutation types, etc.
Supplementary~file~\ref{suppfile:code} also contains text files and CSV files with all of the numerical values shown in these plots.

\subsection{Alignments and phylogenetic analyses of HIV-1 {\it env} sequences}
We made the trees in Fig \ref{ref:ch3_tree} using multiple-sequence alignments downloaded from the HIV sequence database (\url{http://www.hiv.lanl.gov/}) that we then manually curated in the following ways:
First, we removed sequences that differed in length from the HXB2 reference sequence (including gap characters), or which contained a premature stop codon, ambiguous residue, or frame-shift mutation.
Next, we removed columns in the alignment for which we lack deep mutational scanning data or are not present in HXB2, columns that have $>$5\% gap characters, or columns in variable loops that looked poorly aligned by eye.
The sequences used to make the trees in Fig \ref{ref:ch3_tree} were randomly selected from the larger alignment, with a defined number of sequences per clade, as described in the figure.
The topologies of these trees were inferred by \texttt{RAxML} using the GTRCAT model.
In the \texttt{phydms}~\cite{hilton2017phydms} analyses, we fixed the topologies of these trees, and then compared the ability of the different models to optimize the branch lengths.
The trees shown in Fig \ref{ref:ch3_tree} are the results of optimizing the branch lengths using the YNGKP M5 model.
The code we used to generate these alignments, build the trees, and conduct the \texttt{phydms} analyses will be made available upon publication of this work.

\subsection{Preference differences}Next, we sought to quantify shifts in Env's preferences at a site-specific level.
Correlating preferences between replicates of the same homolog showed that some sites were substantially influenced by experimental noise Fig \ref{fig:ch3_homologs_corr}.
Thus, we quantified shifts using an approach that corrects for noise, similar to the one used in a previous study that deep mutational-scanning data between homologs of influenza's nucleoprotein \cite{doud2015site}.
Fig \ref{fig:ch3_distances}A shows how we quantified shifts for a few example sites.
For a given site $r$, we define the distance in preferences between an arbitrary pair of replicates $i$ and $j$ as: $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.
We then use this metric to measure the distance between all pairwise combinations of replicates from both homologs.
We quantified the un-corrected distance between homologs as the root mean square of all replicate-replicate distances between homologs ($RMSD_{between}$).
Since $D_{r}^{i,j}$ can range between 0-1, $RMSD_{between}$ also ranges between 0-1, with 0 indicating no differences and 1 indicating extreme differences between the homologs.
Next, we quantified experimental noise as the root mean square of all replicate-replicate distances where both replicates came from the \textit{same} homolog ($RMSD_{within}$).
Finally, we computed the noise-corrected distance between homologs ($RMSD_{corrected}$) by subtracting $RMSD_{between}$ by $RMSD_{within}$.

Sites where we repeatedly measured large differences between homologs are expected to have $RMSD_{corrected}$ values much greater than zero (e.g., sites 512 and 309; Fig \ref{fig:ch3_distances}A).
In contrast, sites where we repeatedly measured small differences between homologs are expected to have $RMSD_{corrected}$ values close to zero (e.g., site 296).
Sites where the noise completely overwhelmed the biological signal are also expected to have $RMSD_{corrected}$ values close to zero (e.g., site 607).
Thus, a $RMSD_{corrected}$ value near zero could indicate that a site's preferences are conserved between homologs, or that the noise at a site is high.

\section{Acknowledgments}
We thank Michael Doud and Orr Ashenberg for computer code that formed the basis for some of the analyses.
We thank Andrew Ward for pointing out to us that some of the sites with slower-than-expected rates of evolution are involved in Env's conformational changes upon receptor binding.
We thank Kelly Lee for helpful input about the relative stabilities of the BG505 and BF520 Envs.
We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
\jdbcomment{add more}

>>>>

\begin{figure}
\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.68\textwidth]{figures/entrenchment.pdf}}
\caption{\label{fig:entrenchment}
Entrenchment of substitutions during Env evolution.
There are 12,521 possible amino-acid mutations at the 659 mutagenized sites alignable between BG505 and BF520.
The blue densities show the effects of all these mutations to each Env.
The orange densities show the effects of just the 92 mutations that convert BG505 to BF520 or vice versa.
In the absence of entrenchment, mutating a site in BG505 to its identity in BF520 should have the opposite effect of mutating the site in BF520 to its identity in BG505.
In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution -- and both distributions should be centered around zero if the two Envs are equivalently functional.
Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions are often entrenched in the Env in which they have fixed.
The effect of a mutation is quantified as the log of the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
}
\end{figure}

@@

The end primers for the BG505 mutagenesis were: \seqsplit{5'-tgaaggcaaaactactggtccgtctcgagcagaagacagtggcaatgaga-3'} and \seqsplit{5'-gctacaaatgcatataacagcgtctcattctttccctaacctcaggcca-3'}.
As with BF520, we cloned the BG505 \textit{env} libraries into the \textit{env} locus of the full-length proviral genome of HIV strain Q23 \citep[another subtype-A transmitted/founder virus;][]{poss1999variants} using the high-efficiency cloning vector described in \citet{dingens2017comprehensive}.
For this cloning, we digested the cloning vector with BsmBI, and then used PCR to elongate the amplicons to include 30 nucleotides at each end that were identical in sequence to the ends of the BsmBI-digested vector.
The primers for this PCR were:
\seqsplit{5'-agataggttaattgagagaataagagaaagagcagaagacagtggcaatgagagtgatgg-3'} and \seqsplit{5'-ctcctggtgctgctggaggggcacgtctcattctttccctaacctcaggccatcc-3'}.
Next, we used NEBuilder HiFi DNA Assembly (NEB, E2621S) to clone the \textit{env} amplicons into the BsmBI-digested plasmids.
We purified the assembled products using Agencourt AMPure XP beads (Beckman Coulter, A63880) using a bead-to-sample ratio of 1.5, and then transformed the purified products into Stellar electrocompetent cells (Takara, 636765).
The transformations yielded between 1.5-3.6 million unique clones for each of the three replicate libraries, as estimated by plating 1:2,000 dilutions of the transformations.
We scraped the plated colonies and maxiprepped the plasmid DNA; unlike in \citet{dingens2017comprehensive}, we did not include a 4-hour outgrowth step after the scraping step.
For the wildtype controls, we maxiprepped three independent cultures of wildtype BG505 \textit{env} cloned into the same Q23 proviral plasmid.
See Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger} and Figure~\ref{fig:mutfreqs}A for information on the average mutation rate in these libraries as estimated by Sanger sequencing and deep sequencing, respectively.

@@

\subsection{Analysis of deep-sequencing data}
We analyzed the deep-sequencing data using the \texttt{dms\_tools2} software package~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.
The algorithm that goes from the deep-sequencing counts to the amino-acid preferences is that described in \citet{bloom2015software} (see also \url{https://jbloomlab.github.io/dms_tools2/prefs.html}).
A Jupyter notebook that performs the entire analysis including generation of most of the figures in this paper is in Supplementary~file~\ref{suppfile:code}.
An HTML rendering of this notebook is in Supplementary~file~\ref{suppfile:html}.

The Jupyter notebooks in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} also contain numerous plots that summarize relevant aspects of the deep sequencing such as read depth, per-codon mutation frequency, mutation types, etc.
Supplementary~file~\ref{suppfile:code} also contains text files and CSV files with the numerical values shown in these plots.

Citations are also owed to \texttt{weblogo}~\citep[][\url{http://weblogo.threeplusone.com/}]{crooks2004weblogo} and \texttt{ggseqlogo}~\citep[][\url{https://omarwagih.github.io/ggseqlogo/}]{wagih2017ggseqlogo}, which were used in the generation of the logoplots.

\subsection{Alignments and phylogenetic analyses of Env sequences}
A basic description of the process used to generate the clade A sequence alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, the alignment mask in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, and the phylogenetic tree in Figure~\ref{fig:tree} are provided in the legend to that figure.
An algorithmic description of how the alignment and tree were generated are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.

For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in Figure~\ref{fig:tree}.
The Goldman-Yang (or YNGKP) model used in Table~\ref{tab:phydms} is the M5 variant described by \citet{yang2000codon}, with the equilibrium codon frequencies determined empirically using the CF3x4 method~\citep{pond2010correcting}.
For the ExpCM shown in Table~\ref{tab:phydms}, we extended the models with empirical nucleotide frequencies described in \citet{hilton2017phydms} to also allow $\omega$ to be drawn from discrete gamma-distributed categories exactly as for the M5 model.
These ExpCM with gamma-distributed $\omega$ were implemented in \texttt{phydms} using the equations provided by \citet{yang1994maximum} (see also \url{http://jbloomlab.github.io/phydms/implementation.html#models-with-a-gamma-distributed-model-parameter}).
For both the M5 model and the ExpCM with a gamma-distributed $\omega$, the number of categories in the discretized gamma distribution is four.

Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for a wider set of models than those used in Table~\ref{tab:phydms}. 
These include the M0 model of \citet{yang2000codon}, ExpCM without a gamma-distributed $\omega$, and ExpCM in which the amino-acid preferences are averaged across sites as a control to ensure that the improved performance of these models is due to their site-specificity.
Note how for these Env alignments, using a gamma-distributed $\omega$ is very important in order for the ExpCMs to outperform the M5 model -- we suspect this is because there are many sites of strong diversifying selection.

For detection of sites with faster or slower than expected evolution, we used the approach in \citet{bloom2017identification}, which is exactly modeled on the FEL approach of \citet{kosakovsky2005not} but extended to ExpCM.
This approach estimates a $P$-value that $\omega_r$ is not equal to one for each site $r$ using a likelihood-ratio test.
For the Q-values and false discovery rate testing, we considered the tests for $\omega_r > 1$ and $\omega_r < 1$ separately.

Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contains the code that runs \texttt{phydms} to reproduce all of these analyses.

\subsection{Identifying sites of shifted amino-acid preference}
When identifying shifts in amino-acid preferences between the two Envs, we needed a way to quantify differences between the Envs while accounting for the fact that our measurements are noisy.
The approach we use is based closely on that of \citet{doud2015site}, and is illustrated graphically in Figure~\ref{fig:prefsdist}A.
The RMSD$_{\rm{corrected}}$ value is our measure of the magnitude of the shift.
Figure~\ref{fig:prefsdist}A, its legend, and the associated text completely explains these calculations with the following exception: they do not detail how the ``distance'' between any two preference measurements was calculated. 
The distance between preferences at each site was simply defined as half of the sum of absolute value of the difference between preferences for each amino acid.
Specifically, for a given site $r$, let $\pi_{r,a}^{i}$ be the preference for amino-acid $a$ in homolog $i$ (e.g., BG505) and let $\pi_{r,a}^{j}$ be the preference for that same amino acid in homolog $j$ (e.g., BF520). 
Then the distance between the homologs at this site is simply $D_{r}^{i,j} = \frac{1}{2}\sum_{a}|\pi_{r,a}^{i}-\pi_{r,a}^{j}|$.
The factor of $\frac{1}{2}$ is used so that the maximum distance will always fall between zero and one.

\subsection{Analysis of entrenchment}
For the analysis in Figure~\ref{fig:entrenchment}, the results are presented in terms of the mutational effects rather than the amino-acid preferences. 
If $\pi_{r,a}$ is the preference of site $r$ for amino-acid $a$ and $\pi_{r,a'}$ is the preference for amino-acid $a'$ (both re-scaled by the stringency parameters in Table~\ref{tab:phydms}), then estimated effect of the mutation from $a$ to $a'$ is simply $\log\left(\frac{\pi_{r,a'}}{\pi_{r,a}}\right)$.

\subsection{Data and code availability}
All code and input data required to reproduce all analyses in this paper are in Supplementary~file~\ref{suppfile:code} (see also Supplementary~file~\ref{suppfile:html}).
The deep sequencing data are on the Sequence Read Archive with the accession numbers listed in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.

\section{Acknowledgments}
We thank Michael Doud and Orr Ashenberg for computer code that formed the basis for some of the analyses.
We thank Andrew Ward for pointing out to us that some of the sites with slower-than-expected rates of evolution are involved in Env's conformational changes upon receptor binding.
We thank Kelly Lee for helpful input about the relative stabilities of the BG505 and BF520 Envs.
We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
\jdbcomment{Add grants for \textit{bioRxiv} but not \textit{eLife}.}
